Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
I think swapping those two paragraphs improves matters
grr. I tend to confuse i.e. and e.g.

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage[utf8]{inputenc}

@@

@@

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
languages that currently exist, PyPy's tracing JIT operates ``one level down'',
e.g. it traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to arrive at a JIT compiler that can be
applied to a variety of dynamic languages, given an appropriate interpreter for each of them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but has already produced some
promising results, which we will discuss in Section \ref{sect:evaluation}.

@@

It is important to understand how the tracer recognizes that the trace it
recorded so far corresponds to a loop.
This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
e.g. usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, e.g. a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: during
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code executed; during tracing
they are the place where the check for a closed loop is performed.

@@

The tracer interprets these functions in a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. However, they correspond to different
paths through the loop and different ways to unroll it. To ascertain which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be executed. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

@@

The simple insight how to improve the situation is that most of the operations
in the trace are actually concerned with manipulating the bytecode string and
the program counter. Those are stored in variables that are ``green'' (e.g. they
are part of the position key).  This means that the tracer checks that those
variables have some fixed value at the beginning of the loop (they may well
change over the course of the loop, though). In the example of Figure
\ref{fig:trace-no-green-folding} the check would be that at the beginning of the
trace the \texttt{pc} variable is \texttt{4} and the \texttt{bytecode} variable
is the bytecode string corresponding to the square function. Therefore it is
possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

From then on things proceed like described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it finds one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (e.g. this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation. Usually this means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter. After the language interpreter takes over again, the whole process
starts again.

@@

\textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function considerably,
making it about six times faster than the pure interpreter.

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked. In this way the overhead of the profiling is measured. For this interpreter
it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
likely that the overhead is less significant.

\begin{figure}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
5 &Profile Overhead &515.0 $\pm$ 7.2 &0.86\\
\hline
\end{tabular}
\label{fig:bench1}
\caption{Benchmark results of example interpreter computing the square of
10000000}
\end{figure}

Furthermore, to test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-example}. The function is a bit arbitrary but executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations (like indexing into the tuple, addition
and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
Python).

@@

\begin{figure}
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.4 &4.96 $\pm$ 0.05 &4.73\\
\hline
\end{tabular}
\caption{Benchmarked function and results for the Python interpreter running
\texttt{f(10000000)}}
\end{figure}

@@

Applying a trace-based optimizer to an interpreter and adding hints to help the
tracer produce better results has been tried before in the context of the DynamoRIO
project \cite{sullivan_dynamic_2003}, which has been a great inspiration for our
work. They achieve the same unrolling of the interpreter loop so that the
unrolled version corresponds to the loops in the user program. However the
approach is greatly hindered by the fact that they trace on the machine code
level and thus have no high-level information available about the interpreter.
This makes it necessary to add quite a large number of hints, because at the
assembler level it is not really visible anymore that e.g. a bytecode string is
immutable. Also more advanced optimizations like allocation removal would
not be possible with that approach.

@@

\textbf{Optimizing Frame Objects:} One problem with the removal of allocations
is that many dynamic languages are so reflective that they allow the
introspection of the frame object that the interpreter uses to store local
variables (e.g. SmallTalk, Python). This means that intermediate results always
escape because they are stored into the frame object, rendering the allocation
removal optimization ineffective. To remedy this problem we make it possible to
update the frame object lazily only when it is actually accessed from outside of
the code generated by the JIT.

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage[utf8]{inputenc}

@@

\newcommand\ie{i.e.,\xspace}
\newcommand\eg{e.g.,\xspace}

@@

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
languages that currently exist, PyPy's tracing JIT operates ``one level down'',
\ie it traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to arrive at a JIT compiler that can be
applied to a variety of dynamic languages, given an appropriate interpreter for each of them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but has already produced some
promising results, which we will discuss in Section \ref{sect:evaluation}.

@@

It is important to understand how the tracer recognizes that the trace it
recorded so far corresponds to a loop.
This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
\ie usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, \eg a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: during
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code executed; during tracing
they are the place where the check for a closed loop is performed.

@@

The tracer interprets these functions in a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (\eg the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

@@

The condition for reusing already existing machine code also needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. However, they correspond to different
paths through the loop and different ways to unroll it. To ascertain which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be executed. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

The simple insight how to improve the situation is that most of the operations
in the trace are actually concerned with manipulating the bytecode string and
the program counter. Those are stored in variables that are ``green'' (\ie they
are part of the position key).  This means that the tracer checks that those
variables have some fixed value at the beginning of the loop (they may well
change over the course of the loop, though). In the example of Figure
\ref{fig:trace-no-green-folding} the check would be that at the beginning of the
trace the \texttt{pc} variable is \texttt{4} and the \texttt{bytecode} variable
is the bytecode string corresponding to the square function. Therefore it is
possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

From then on things proceed like described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it finds one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (\eg this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation. Usually this means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter. After the language interpreter takes over again, the whole process
starts again.

@@

\textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function considerably,
making it nearly three times faster than the pure interpreter.

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked. In this way the overhead of the profiling is measured. For this interpreter
it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (\eg the Python one) it seems 
likely that the overhead is less significant.

\begin{figure}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
5 &Profile Overhead &515.0 $\pm$ 7.2 &0.86\\
\hline
\end{tabular}
\caption{Benchmark results of example interpreter computing the square of
\label{fig:bench1}
10000000}
\end{figure}

Furthermore, to test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. The function is a bit arbitrary but executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations (like indexing into the tuple, addition
and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
Python).

@@

\begin{figure}
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.4 &4.96 $\pm$ 0.05 &4.73\\
\hline
\end{tabular}

\caption{Benchmarked function and results for the Python interpreter running
\texttt{f(10000000)} \label{fig:bench-python}}
\end{figure}

@@

Applying a trace-based optimizer to an interpreter and adding hints to help the
tracer produce better results has been tried before in the context of the DynamoRIO
project \cite{sullivan_dynamic_2003}, which has been a great inspiration for our
work. They achieve the same unrolling of the interpreter loop so that the
unrolled version corresponds to the loops in the user program. However the
approach is greatly hindered by the fact that they trace on the machine code
level and thus have no high-level information available about the interpreter.
This makes it necessary to add quite a large number of hints, because at the
assembler level it is not really visible anymore that \eg a bytecode string is
immutable. Also more advanced optimizations like allocation removal would
not be possible with that approach.

@@

\textbf{Optimizing Frame Objects:} One problem with the removal of allocations
is that many dynamic languages are so reflective that they allow the
introspection of the frame object that the interpreter uses to store local
variables (\eg SmallTalk, Python). This means that intermediate results always
escape because they are stored into the frame object, rendering the allocation
removal optimization ineffective. To remedy this problem we make it possible to
update the frame object lazily only when it is actually accessed from outside of
the code generated by the JIT.
